### 2016-02-12 prepReport/theorBack.tex
Implemented half of the suggestions of JCF

\subsection{Basic Concepts of Machine Learning}\label{sec:tb_basicCon}
Machine Learning is a field of Computer Science that studies the development of mathematical techniques that allow software to learn autonomously, without an explicit description of each rule of operation. Its goal is to extract latent features from the data that allow an immediate classification of each input data into a particular class -- the catch is that there is no previous rule formulation, but instead we have an adaptive model that adjusts is parameters accordingly with the input data it receives, improving the estimates it yields as it receives new input samples.

Let us consider a practical example. For instance, suppose we want to build a program that given an input audio waveform representation of a spoken word, it matches it into a particular word of a dictionary. We could, of course, devise a set of rules and exceptions for each word analysing some of its features (perhaps the Fourier representation of each one, and, from it, manually finding the appropriate rules for each), but apart from it being a very complex task, it wouldn't be a scalable solution, given the enormous number of words in each language. The approach taken by Machine Learning is different, and instead of manually processing each waveform, we build a large dataset, of size $N$, containing the waveforms of several words $\left[ \mb{x}_1 \; \mb{x}_2 \;\mb{x}_2 \; \dots \mb{x}_N \right]$ -- we call this dataset the \textbf{training dataset} -- and we feed it to our model. Each of the $i$-th data point was previously labelled, and in fact we feed each training data point $\mb{x}_i$ \emph{along} with its corresponding label $t_i$, so that the model can adapt its parameters accordingly to the \emph{target value} it is supposed to classify. This set of labels $\mb{t} = \left[ t_1 \; t_2 \; t_3 \; \dots \; t_N \right]$ is called the \textbf{target vector}.

@@

Now that the basic Machine Learning concepts have been presented, I will discuss, henceforth, one of the most important algorithms that address the supervised classification problems, the \emph{Artificial Neural Networks} that will be discussed in Section~\ref{sec:theorBack_ann} as somewhat of a contextual introduction to the main theme of the thesis, which will be Recurrent Neural Networks (Section~\ref{sec:theorBack_rnn}), namely the \textbf{Long Short-Term Memory Networks} (Section~\ref{sec:theorBack_lstm}), both of which are improvements over the initial formulation of the ANNs. These two last networks branch even further from these set of problems, and are usually employed in \textit{Deep Learning} tasks, where we try to extract even higher level information from data at the expense of increased model complexity.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.9\linewidth]{figures/activFunc.eps}
    \caption[Possible activation functions of choice for a neuron]{Three different activation functions. As you can see, the hyperbolic tangent has the same extreme value as the sign step function, but has a smooth transition between them, which can be interpreted as a \emph{soft decision} in the more ambiguous middle region, reflecting the degree of uncertainty on the decision. On the other hand, the sigmoid function goes from zero to one, and is also smooth like the hyperbolic tangent}.
	\label{fig:activFunc}
\end{figure}

A neuron by itself can be thought of as a simple linear regression, where we optimize the weight of each feature according to a target value, or function. While important in some applications, the main interest in ANN is to evaluate increasingly more complex models, and not a simple linear regression. This is achieved by \emph{chaining} nodes to one another, connecting the output of a given node to one of the inputs of another. We call \emph{layers} to a group of these nodes that occupy the same hierarchical position. There can be any number of layers, with any number of nodes, but most implementations generally have 3 layers: the \emph{initial} layer, the \emph{hidden} layer (in the middle) and the \emph{output} layer. Figure~\ref{fig:neuralnet} suggests a possible structure for a 3 layer ANN.

@@

\begin{equation}\label{eq:deltaj}
	\delta_j = f'(a_j) \sum_k w_{kj} \delta_k .
\end{equation}
The interpretation of these equations is simple. If $w_{ji}$ is the weight of the connection between the neuron $j$ we are considering and a neuron $i$ in a previous layer, then the sum over $k$ relates to all the neurons in the \emph{next} layer to which $j$ connects: this way, since the update on $w_{ji}$, according to~\ref{eq:gradDesc}, is given by

@@

The training of an RNN is usually performed using a variation of Backpropagation, called \textbf{Backpropagation Through Time} (BPTT), that as the name implies, performs the same backpropagation procedure discussed for the ANNs, but now taking into account the unfolding of the network through a fixed training epoch $T$ like Figure~\ref{fig:recneuron_unf}. Due to this very fact, this training procedure is memory and performance consuming, and so it will not be used in my final work, but instead a novel approach, the \textbf{Simultaneous Perturbation Stochastic Approximation}.

@@

\subsection{Long Short-Term Memory Networks}\label{sec:theorBack_lstm}
To overcome the issue of failing to remember long-term dependencies, Hochreiter and Schmidhüber proposed, in 1997, a novel approach to the RNNs called the \textit{Long-Short Term Network}~\cite{Hoch97}. This section explains the main idea of this approach~\ref{sec:struct_lstm}, and also how it is trained\ref{sec:training_lstm}, serving as a support for the work of this thesis.
Although originally formulated in 1997, its formulation has been incrementally updated in~\cite{Gers00} and~\cite{Gers2000}, and the most current version is the one in~\cite{Graves05}. One of the inital proposers of LSTM, Prof. Jürgen Schmidhüber, did a survey on the most common variations of the model last year~\cite{Greff15}, and this will the basis of this short theoretical presenation, as well as the work that will be developed in this thesis.

\subsubsection{Structure, Operation and Equations}\label{sec:struct_lstm}
A single LSTM neuron is presented in Figure~\ref{fig:lstmneuron}. As we can see from the picture, we still have the recurrent connections from the regular RNNs, but now there are multiple entry points that control the flow of information through the network. Although ommited from the  picture, all the gates are biased, as is suggested in Equations~\ref{eq:equationsLstm}. The main components, their role and relevance, are explained as follows

@@

    \item \textbf{Block Input Gate} -- as the name implies, this gate controls the flow of information from the input gate to the memory cell. It also receives the input vector and the previous output vector as inputs, but it does not have peephole connections and its dynamics are controlled by a different set of weights. The \textbf{activation function of this gate can be any}, but the most common choice is the \textbf{Hyperbolic Tangent}.

@@

    \item \textbf{Memory Cell} -- the cornerstone of the LSTM neuron. This is the memory element of the neuron, where the previous state is kept, and updated accordingly to the dynamics of the gates that connect to it. Also, this is where the peephole connections come from:

@@

    \item \textbf{Peepholes} -- direct connections \textit{from} the memory cell that allow for gates to `peep' the states of the memory cell. They were added after the initial 1997 formulation, and their absence was proven to have a minimal performance impact~\cite{Greff15}.
\end{itemize}

@@

\begin{align}\label{eq:equationsLSTM}
    \mb{z}^{(t)} & = g(\mb{W}_z \mb{x}^{(t)} + \mb{R}_z \mb{y}^{(t-1)} + \mb{b}_z) \\
    \mb{i}^{(t)} & = \sigma(\mb{W}_i \mb{x}^{(t)} + \mb{R}_i \mb{y}^{(t-1)} + \mb{p}_i \odot \mb{c}^{(t-1)} + \mb{b}_i) \\
    \mb{f}^{(t)} & = \sigma(\mb{W}_f \mb{x}^{(t)} + \mb{R}_f \mb{y}^{(t-1)} + \mb{p}_f \odot \mb{c}^{(t-1)} + \mb{b}_f) \\
    \mb{c}^{(t)} & = \mb{i}^{(t)} \odot \mb{z}^{(t)} + \mb{f}^{(t)} \odot \mb{c}^{(t-1)} \\
    \mb{o}^{(t)} & = \sigma(\mb{W}_o \mb{x}^{(t)} + \mb{R}_o \mb{y}^{(t-1)} + \mb{p}_o \odot \mb{c}^{(t)} + \mb{b}_o) \\
    \mb{c}^{(t)} & = \mb{o}^{(t)} \odot h(\mb{z}^{(t)}) 
\end{align}
The $i$-th element of the previous vectors in bold corresponds to the value of the gate of the $i$-th neuron of the layer, which is a very convenient and compact representation of the whole layer. Furthermore, if the layer has $N$ LSTM neurons and $M$ inputs (i.e. the size of the layer that precedes this), we see that the input weight matrices $\mb{W}_*$ and the  recurrent weight matrices $\mb{R}_*$ all have size $N \times M$, and that the bias weight matrices $\mb{b}_*$, the peephole weight matrices $\mb{p}_*$ and the matrices $\mb{z}^{(t)}$ through $\mb{c}^{(t)}$ have size $N \times 1$. Although useful for a high-level description in a programming language, the matrix representation  may not be suitable for a direct HDL port. In that case, in order to represent the operation of the $i$-\textit{single neuron}, all the above equations still hold, but instead of \textit{vectors} of outputs, we will have a single scalar output, and we only use the appropriate row in the weight matrices $\mb{W}_{*}$, $\mb{R}_{*}$ and the remaining.

\subsubsection{Training -- SPSA}\label{sec:training_lstm}
Since this work aims to perform on-chip learning, it is important to find a suitable learning scheme. Since I am aiming at an hardware implementation, and although the memory resources of nowadays FPGAs are abundant, one must find an algorithm that uses the less memory as possible (at the smallest performance cost possible), in order to use the additional memory, for instance, to add additional LSTM cells that can improve the performance of our system. 
That being said, we see that the usual training algorithms for LSTM cells -- i.e. Real-Time Recurrent Learning (RTRL), Truncated Backpropagation Through Time (BPTT) or a mixture of both~\cite{Graves05}~\cite{Hoch97}~\cite{Greff15} -- usually involve the storage of the deltas of each layer for every time instant in the training epoch (from $0$ to $T$), which is a highly non-scalable solution both in terms of memory and performance. A most efficient approach to training times series dependant structures like LSTM is the use of \textbf{Simultaneous Perturbation Stochastic Approximation} (SPSA)~\cite{Spall98}.  The main idea of this technique is, instead of explicitly evaluate the gradients for the cost function at each time step, perform a random walk in the neighbourhood of the current weight matrix, in the weight space, and approximate the new weight update by the approximation of the gradient of the cost function that resulted from that random walk. The weight update for the $i$-th weight at the time step $t$ is given by

\begin{equation}\label{eq:spsa_weightUpdate}
    \Delta w_i^{(t)} = \frac{J(\mb{w}^{(t)} + \beta \mb{s}^{(t)}) - J(\mb{w}^{(t)})}{\beta s_i^{(t)}}
\end{equation}
where $\beta$ is the magnitude of the perturbation to be introduced, $\mb{s}^{(t)}$ is a \textit{sign vector} whose $i$-th element, $s_i^{(t)}$, is either $-1$ or $1$. This way, we see that every weight is \textbf{randomly} incremented either by $-\beta$ or $\beta$, and we only need to keep a duplicate of the weight matrix with the perturbations, and we only need to evaluate the cost function twice per incoming sample. As for the \textit{update rule} is concerned, we have

@@

\section{Proposed Solution}\label{sec:propSol}
Given the previous discussion of how LSTM neurons work and their benefits when compared to other RNN architectures, this work aims to conceive an efficient dedicated FPGA implementation with on-chip learning that, hopefully, surpasses the training/testing time of an equivalent CPU implementation in a general high-level language or of a framework.

As discussed in Chapter~\ref{chap:stateArt}, hitherto there is only one FPGA implementation of LSTM cells~\cite{Chang15}, but there is no on-chip learning on the architecture (the weights are learned offline, and loaded in the FPGA before operation, and they are not changed while the FPGA system is operating), and my proposed work will implement an efficient LSTM implementation \emph{with} the ability to learn on-chip, which is an innovative and useful feature for the LSTM architecture on a dedicated hardware implementation. I will achieve that using the SPSA method described in~\ref{sec:training_lstm}, and inspired in the previous work of~\cite{Maeda05}.

>>>>

\subsection{Basic Concepts of Machine Learning}\label{sec:tb_basicCon}
Machine Learning is a field of Computer Science that studies the development of mathematical techniques that allow software to learn autonomously, without an explicit description of each rule of operation. Its goal is to extract latent features from the data that allow an immediate classification of each input data into a particular class -- the catch is that there is no previous rule formulation, but instead we have an adaptive model that adjusts is parameters according to the input data it receives, improving the estimates it yields as it receives new input samples.

Let us consider a practical example. For instance, suppose we want to build a program that given an input audio waveform representation of a spoken word, it matches it into a particular word of a dictionary. We could, of course, devise a set of rules and exceptions for each word analysing some of its features (perhaps the Fourier representation of each one, and, from it, manually finding the appropriate rules for each), but apart from being a very complex task, it wouldn't be a scalable solution, given the enormous number of words in each language. The approach taken by Machine Learning is different, and instead of manually processing each waveform, we build a large dataset, of size $N$, containing the waveforms of several words $\left[ \mb{x}_1 ,\; \mb{x}_2 ,\;\mb{x}_2 ,\; \dots ,\;\mb{x}_N \right]$ -- we call this dataset the \textbf{training dataset} -- and we feed it to our model. Each of the $i$-th data point was previously labelled, and in fact we feed each training data point $\mb{x}_i$ \emph{along} with its corresponding label $t_i$, so that the model can adapt its parameters accordingly to the \emph{target value} it is supposed to classify. This set of labels $\mb{t} = \left[ t_1 ,\; t_2 ,\; t_3 ,\; \dots ,\; t_N \right]$ is called the \textbf{target vector}.

@@

Now that the basic Machine Learning concepts have been presented, I will discuss, henceforth, one of the most important algorithms that address the supervised classification problems, the \emph{Artificial Neural Networks} that will be discussed in Section~\ref{sec:theorBack_ann} as a contextual introduction to the main theme of the thesis, which will be Recurrent Neural Networks (Section~\ref{sec:theorBack_rnn}), namely the \textbf{Long Short-Term Memory Networks} (Section~\ref{sec:theorBack_lstm}), both of which are improvements over the initial formulation of the ANNs. These two last networks branch even further from these set of problems, and are usually employed in \textit{Deep Learning} tasks, where we try to extract even higher level information from data at the expense of increased model complexity.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.9\linewidth]{figures/activFunc.eps}
    \caption[Possible activation functions of choice for a neuron]{Three different activation functions. As you can see, the hyperbolic tangent has the same extreme values as the sign step function, but has a smooth transition between them, which can be interpreted as a \emph{soft decision} in the more ambiguous middle region, reflecting the degree of uncertainty on the decision. On the other hand, the sigmoid function goes from zero to one, and is also smooth like the hyperbolic tangent.}
	\label{fig:activFunc}
\end{figure}

A neuron by itself can be thought of as a simple linear regression, where we optimize the weight of each feature according to a target value, or function. While important in some applications, the main interest in ANN is to evaluate increasingly more complex models, and not a simple linear regression. This is achieved by \emph{chaining} nodes to one another, connecting the output of a given node to one of the inputs of another. We call \emph{layers} to a group of these nodes that occupy the same hierarchical position. There can be any number of layers, with any number of nodes, but most implementations generally have 3 layers: the \emph{input} layer, the \emph{hidden} layer (in the middle) and the \emph{output} layer. Figure~\ref{fig:neuralnet} suggests a possible structure for a 3 layer ANN.

@@

\begin{equation}\label{eq:deltaj}
	\delta_j = f'(a_j) \sum_k w_{kj} \delta_k .
\end{equation}
The interpretation of these equations is simple. If $w_{ji}$ is the weight of the connection between the neuron $j$ we are considering and a neuron $i$ in a previous layer, then the sum over $k$ relates to all the neurons in the \emph{next} layer to which $j$ connects: this way, since the update of $w_{ji}$, according to~\ref{eq:gradDesc}, is given by

@@

The training of an RNN is usually performed using a variation of Backpropagation, called \textbf{Backpropagation Through Time} (BPTT), that as the name implies, performs the same backpropagation procedure discussed for the ANNs, but now taking into account the unfolding of the network through a fixed training epoch $T$ like Figure~\ref{fig:recneuron_unf}. Due to this very fact, this training procedure is memory and performance consuming, and so it will not be used in my final work, but instead a novel approach, the \textbf{Simultaneous Perturbation Stochastic Approximation} will be evaluated.

@@

\subsection{Long Short-Term Memory Networks}\label{sec:theorBack_lstm}
To overcome the issue of failing to remember long-term dependencies, Hochreiter and Schmidhüber proposed, in 1997, a novel approach to the RNNs called the \textit{Long-Short Term Network}~\cite{Hoch97}. This section explains the main idea of this approach (Section ~\ref{sec:struct_lstm}), and also how it is trained (Section~\ref{sec:training_lstm}), serving as a support for the work of this thesis.
Although originally formulated in 1997, its formulation has been incrementally updated in~\cite{Gers00} and~\cite{Gers2000}, and the most current version is the one in~\cite{Graves05}. One of the inital proposers of LSTM, Prof. Jürgen Schmidhüber, did a survey on the most common variations of the model last year~\cite{Greff15}, and this will the basis of this short theoretical presentation, as well as the work that will be developed in this thesis.

\subsubsection{Structure, Operation and Equations}\label{sec:struct_lstm}
A single LSTM neuron is presented in Figure~\ref{fig:lstmneuron}. As we can see from the picture, we still have the recurrent connections from the regular RNNs, but now there are multiple entry points that control the flow of information through the network. Although omitted from the picture, all the gates are biased, as is suggested in Equations~\ref{eq:equationsLstm}. The main components, their role and relevance, are explained as follows

@@

    \item \textbf{Block Input Gate} -- as the name implies, this gate controls the flow of information from the input gate to the memory cell. It also receives the input vector and the previous output vector as inputs, but it does not have peephole connections and its dynamics are controlled by a different set of weights. The \textbf{activation function of this gate can be either}, but the most common choice is the \textbf{Hyperbolic Tangent}.

@@

    \item \textbf{Memory Cell} -- the cornerstone of the LSTM neuron. This is the memory element of the neuron, where the previous state is kept, and updated accordingly to the dynamics of the gates that connect to it. Also, this is where the peephole connections come from.

@@

    \item \textbf{Peepholes} -- direct connections \textit{from} the memory cell that allow for gates to `peep' at the states of the memory cell. They were added after the initial 1997 formulation, and their absence was proven to have a minimal performance impact~\cite{Greff15}.
\end{itemize}

@@

\begin{align}\label{eq:equationsLSTM}
    \mb{z}^{(t)} & = g(\mb{W}_z \mb{x}^{(t)} + \mb{R}_z \mb{y}^{(t-1)} + \mb{b}_z) \\
    \mb{i}^{(t)} & = \sigma(\mb{W}_i \mb{x}^{(t)} + \mb{R}_i \mb{y}^{(t-1)} + \mb{p}_i \odot \mb{c}^{(t-1)} + \mb{b}_i) \\
    \mb{f}^{(t)} & = \sigma(\mb{W}_f \mb{x}^{(t)} + \mb{R}_f \mb{y}^{(t-1)} + \mb{p}_f \odot \mb{c}^{(t-1)} + \mb{b}_f) \\
    \mb{c}^{(t)} & = \mb{i}^{(t)} \odot \mb{z}^{(t)} + \mb{f}^{(t)} \odot \mb{c}^{(t-1)} \\
    \mb{o}^{(t)} & = \sigma(\mb{W}_o \mb{x}^{(t)} + \mb{R}_o \mb{y}^{(t-1)} + \mb{p}_o \odot \mb{c}^{(t)} + \mb{b}_o) \\
    \mb{c}^{(t)} & = \mb{o}^{(t)} \odot h(\mb{z}^{(t)}) 
\end{align}
where $\odot$ is the Hadamard multiplication. The $i$-th element of the previous vectors in bold corresponds to the value of the gate of the $i$-th neuron of the layer, which is a very convenient and compact representation of the whole layer. Furthermore, if the layer has $N$ LSTM neurons and $M$ inputs (i.e. the size of the layer that precedes this), we see that the input weight matrices $\mb{W}_*$ and the  recurrent weight matrices $\mb{R}_*$ all have size $N \times M$, and that the bias weight matrices $\mb{b}_*$, the peephole weight matrices $\mb{p}_*$ and the matrices $\mb{z}^{(t)}$ through $\mb{c}^{(t)}$ have size $N \times 1$. Although useful for a high-level description in a programming language, the matrix representation  may not be suitable for a direct HDL port. In that case, in order to represent the operation of the $i$-\textit{single neuron}, all the above equations still hold, but instead of \textit{vectors} of outputs, we will have a single scalar output, and we only use the appropriate row in the weight matrices $\mb{W}_{*}$, $\mb{R}_{*}$ and the remaining.

\subsubsection{Training -- SPSA}\label{sec:training_lstm}
Since this work aims to perform on-chip learning, it is important to find a suitable learning scheme. Since I am aiming at an hardware implementation, and although the memory resources of current FPGAs are abundant, one must find an algorithm that uses the less memory as possible (at the smallest performance cost possible), in order to use the additional memory, for instance, to add additional LSTM cells that can improve the performance of our system. 
That being said, we see that the usual training algorithms for LSTM cells -- i.e. Real-Time Recurrent Learning (RTRL), Truncated Backpropagation Through Time (BPTT) or a mixture of both~\cite{Graves05}~\cite{Hoch97}~\cite{Greff15} -- usually involve the storage of the deltas of each layer for every time instant in the training epoch (from $0$ to $T$), which is a highly non-scalable solution both in terms of memory and performance. A most efficient approach to training times series dependant structures like LSTM is the use of \textbf{Simultaneous Perturbation Stochastic Approximation} (SPSA)~\cite{Spall98}.  The main idea of this technique is, instead of explicitly evaluating the gradients for the cost function at each time step, to perform a random walk in the neighbourhood of the current weight matrix, in the weight space, and approximate the new weight update by the approximation of the gradient of the cost function that resulted from that random walk. The weight update for the $i$-th weight at the time step $t$ is given by

\begin{equation}\label{eq:spsa_weightUpdate}
    \Delta w_i^{(t)} = \frac{J(\mb{w}^{(t)} + \beta \mb{s}^{(t)}) - J(\mb{w}^{(t)})}{\beta s_i^{(t)}}
\end{equation}
where $\beta$ is the magnitude of the perturbation to be introduced, $\mb{s}^{(t)}$ is a \textit{sign vector} whose $i$-th element, $s_i^{(t)}$, is either $-1$ or $1$. This way, we see that every weight is \textbf{randomly} incremented either by $-\beta$ or $\beta$, and we only need to keep a duplicate of the weight matrix with the perturbations, and we only need to evaluate the cost function twice per incoming sample. As for the \textit{update rule}, we have

@@

\section{Proposed Solution}\label{sec:propSol}
Given the previous discussion of how LSTM neurons work and their benefits when compared to other RNN architectures, this work aims to conceive and design an efficient dedicated FPGA implementation with on-chip learning that, surpasses the training/testing time of an equivalent CPU implementation in a general high-level language or of a framework (see Section~\re (see Section~\ref{sec:wp_sh}).

As discussed in Chapter~\ref{chap:stateArt}, up to now there is only one FPGA implementation of LSTM cells~\cite{Chang15}, but there is no on-chip learning on the architecture (the weights are learned offline, and loaded in the FPGA before operation, and they are not changed while the FPGA system is operating), and my proposed work will implement an efficient LSTM implementation \emph{with} the ability to learn on-chip, which is an innovative and useful feature for the LSTM architecture on a dedicated hardware implementation. I will achieve that using the SPSA method described in~\ref{sec:training_lstm}, and inspired in the previous work of~\cite{Maeda05}.
